Follow-up Samples

### 2017-07-22 submission/manuscript_R1.Rmd
add revision and marked up version

***Treatment for colonic lesions alters the bacterial community structure***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1]. None of the individuals in this study had any recorded antibiotic use before treatment of their respective lesion. The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A]. We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05). When analyzing only those who underwent surgery as a treatment these trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A]. Since surgery is another situation where antibiotics are used we tested our $\theta$~YC~ beta diversity in these individuals only. There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma group (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C]. Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value = `r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value = `r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value = `r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].

***The treatment of lesions are not consistent across diagnosis groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. To try and limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs. Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). However, upon randomizing the pre and post treatment labels we found that Random Forest could work equally well (adenoma AUC range = `r format(min(adn_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_randomized_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_randomized_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_randomized_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_randomized_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. Although treatment had an impact on the overall community structure; the effect of treatment was not consistent across patients and diagnosis groups.

>>>>

***Treatment for colonic lesions alters the bacterial community structure***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1]. None of the individuals in this study had any recorded antibiotic usage that was not associated with surgical treatment of their respective lesion. The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A]. We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05). Since surgery is a procedure where antibiotics are used, we tested our $\theta$~YC~ beta diversity in these individuals only. Although not significant the trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A]. There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma group (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C]. Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value = `r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value = `r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value = `r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].

***The treatment of lesions are not consistent across diagnosis groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. To try and limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs. Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. Although treatment had an impact on the overall community structure; the effect of treatment was not consistent across patients and diagnosis groups.
